### 2016-06-15 motivation.tex
Motivation edits.

In this section, we motivate the need for a formal reasoning framework
for weakly isolated transactions through a series of examples. All our
examples are written in C-like imperative language equipped with a
\C{txn} lexical block for transactions. Each \C{txn} block is
associated with a single-quoted string in angle braces that uniquely
identifies the transaction. We use Hoare triple notation to annotate
programs with pre and post conditions.

@@

% Most weak isolation levels allow executions to interleave operations
% (Reads, Writes and Commits) of one transaction with other's, while
% enforcing some additional constraints. Arbitrary interleavings may
% lead to the violation of invaraints. 
Let us consider \iso{Read Committed} (RC), an isolation level weaker
than SER that is default in Postgres 9.5~\cite{postgres95} and Oracle
11g~\cite{oracle11g} databases. An RC transaction is isolated from the
writes of an uncommitted transactions, thus preventing the so called
\emph{dirty reads} phenomenon~\cite{berenson}. In the current example,
RC isolation admits following executions:

@@

An execution is a series of Read, Write and Commit operations.
Subscript of an operation indicates its transaction. For clarity,
effects of different transactions are shown against differently
colored backgrounds. In the first execution, the first transaction
(`Wd1') reads the current balance (\C{k}) and writes the new balance
(\C{k-a1}), but before it commits transaction `Wd2' executes and
commits, writing the new balance (\C{k-a2}). RC isolation has
prevented `Wd2' from witnessing the writes of uncommitted transaction
`Wd1'. Committing `Wd1' now leads to the loss of updates from `Wd2'
(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
in a balance of \C{k-a1}. The second execution describes a similar
scenario with `Wd1' and `Wd2' exchanging their roles.

Clearly, \iso{Read Committed} is not sufficient because it loses the
updates of one transaction and violates the invariant in
Fig.~\ref{fig:motiv-eg-1}. Now, consider an alternative unfolding of
Execution 1, where, rather than committing `Wd1' and overwriting the
updates of `Wd2', we rollback `Wd1' to reexecute it starting from the
state that contains the updates of `Wd2'. No updates are lost in this
alternative execution, leading to the satisfaction of
Fig.~\ref{fig:motiv-eg-1} invariant. Rolling back and reexecuting an
uncommitted transaction to prevent write-write conflicts with a
committed transaction is a strategy often adopted by databases to
implement \iso{Snapshot Isolation} (SI) level. Indeed, SI is
sufficient to enforce the required invaraints in this
example\footnote{\iso{Parallel Snapshot Isolation} (PSI), a weaker
version of SI tailor-made for replicated stores is also sufficient.}.
Moreover, relying on optimistic speculation instead of pessimistic
locking for concurrency control makes SI more efficient than SER,
making it appropriate for transactions in Fig.~\ref{fig:motiv-eg-1}.
Note that we have arrived at his judgment via adhoc reasoning based on
speculations about database's implementation strategies. Such a
reasoning process is necessarily fragile and error-prone. Is there a
better way?

\begin{figure}
\centering
$\{\{\texttt{C+S}\ge\texttt{0}\}\}$
\begin{tabular}{l||l}
\begin{txnimpcode}
  txn$\langle$'WdC'$\rangle${
    if (C+S $\ge$ a1) {
      C := C - a1
    }
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'WdS'$\rangle${
    if (C+S $\ge$ a2) {
      S := S - a2
    }
  }
\end{txnimpcode}
\\
\end{tabular}
$\{\{\texttt{C+S}\ge\texttt{0}\}\}$

\caption{Withdraws from current (\C{C}) and savings (\C{S}) accounts.}
\label{fig:motiv-eg-2}
\end{figure}

\iso{Snapshot Isolation} effectively serializes updates to a data
object without necessarily relying on expensive lock-based concurrency
control. While this makes it a better alternative to
\iso{Serializable} isolation, it is however not a replacement for the
latter. Consider a different banking application that maintains a
current (\C{C}) and savings account (\C{S}) for each user. A user can
withdraw either from current or from savings account as long as the
combined balance is at least as much as the amount being withdrawn.
Fig.~\ref{fig:motiv-eg-2} shows a pair of concurrent transactions,
`WdC' and `WdS', performing withdraws from current and savings
accounts, respectively. Pre and post conditions assert non-negative
combined balance invariant. \iso{Serializable} isolation is sufficient
to preserve the invariant, but \iso{Snapshot Isolation} is not.
Intuitively, this is because there are no \emph{lost updates} in this
example even when both transactions commit concurrently, hence there
is no reason for SI to effectively serialize them. A reasoning
framework for weak isolation should let us formally deduce this fact
by making it impossible to construct a proof of correctness for the
program under SI, but facilitating a proof under SER.

\begin{figure}[t]
\centering
$\{\{\texttt{C+S}\ge\texttt{0}\}\}$
\begin{tabular}{l||l||}
\begin{txnimpcode}
  txn$\langle$'Xfer'$\rangle${
    C := C - a;
    S := S + a;
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  txn$\langle$'TotBal'$\rangle${
    T := S + C
  }
\end{txnimpcode}
% \\
% \multicolumn{2}{c} {
% }
\end{tabular}
\begin{tabular}{c}
  \begin{txnimpcode}
    txn$\langle$'AddInt'$\rangle${
      S := S + (T*0.1)
    }
  \end{txnimpcode}
\end{tabular}

$\{\{\texttt{C+S}\ge\texttt{0}\}\}$

\caption{Concurrent transfer (\C{Xf}) and interest accumulation (\C{Int}).}
\label{fig:motiv-eg-3}
\end{figure}

The previous two examples preserve invariants on \C{C} and \C{S} by
effectively serializing transactions that write to \C{C} and \C{S}.
However, in general, serializing a subset of transactions that write
to the data objects referred by invariants is neither sufficient nor
necessary to uphold the invariants. Consider the banking application
in Fig.~\ref{fig:motiv-eg-3} with three transactions. Transaction
`Xfer' transfers some amount (\C{a}) from current (\C{C}) to savings
(\C{S}) account.  Transaction `TotBal' records the total balance in
both the accounts in a data object \C{T}, and transaction `AddIntr'
calculates interest on total balance (10\% of \C{T}) and adds it to
the savings account. Using a data object to record some function of
the current state of other data objects is a standard practice in
relational databases, where Materialized Views~\cite{oraclematview}
are created to maintain the results of a SQL query. Like the previous
example, the invariant of interest in this example is non-negative
total balance; if we start with a non-negative total balance, and we
add interest based on the total balance, we must end up with a
non-negative total balance.  However, unlike the previous example,
serializing transactions that write to \C{C} and \C{S} is not
sufficient to guarantee the invariant.  The following execution
demonstrates this anomaly:

\begin{figure}[!h]
\centering
\includegraphics[scale=0.5]{Figures/motiv-eg-3-a}

\label{fig:motiv-eg-3-a}
\end{figure}

The subscripts X, TB and AI denote `Xfer', `TotBal' and `AddInt'
transactions, respectively. Assume that \C{C} and \C{S} are initially
10 and 0, respectively. Let \C{a} be 200, i.e., we subtract 200 from
\C{C} and add it to \C{S}.  This operation must be allowed because it
doesn't effect the total balance, hence the interest. The execution
begins with `TotBal' transaction reading the savings (\C{S}) balance
as 0 (we assume left-to-right evaluation order for \C{S + C}).
However, before it reads the current (\C{C}) balance, `Xfer'
transaction executes and commits, changing \C{C} to -190.
Subsequently, `TotBal' reads -190 from \C{C}, and updates \C{T} to
-190. `AddInt', which begins execution after `TotBal', reads -190 from
\C{T}, calculates interest as -19, and updates \C{S} to 181. The
result is a state with total balance as -190 + 181 = -9, thus
violating the invariant.

Note that transactions `Xfer' and `AddInt', which write to the data
objects (\C{C} and \C{S}) relevant to the invariant, are serialized in
the above execution. Transaction `TotBal' however is not. While
serializing it fixes the problem, it is not necessary. A simpler fix
is to isolate `TotBal' transaction from the effects of concurrent
`Xfer' transaction, so that it reads \C{C} as 10 despite the
concurrent `Xfer' transaction committing -190 to \C{C}. This can be
achieved by executing `TotBal' under \iso{Repeatable Read} (RR)
isolation level, which guarantees that any two reads of a data object
in a transaction return the same value (the \emph{repeatable reads}
property~\cite{berenson}). Allowing `Xfer' to interfere with `TotBal'
violates the repeatable read property since a hypothetical read from
\C{S} in `TotBal' after the interference returns a different value
(200) than the first read (0). Executing `TotBal' under RR restores
repeatable read by \emph{effectively} prohibiting such interference.

\begin{figure}[t]
\centering
\hspace*{-0.3in}$\{\{\texttt{V=0}\}\}$
\begin{tabular}{l||l}
\begin{txnimpcode}
  txn$\langle$'Rd'$\rangle${
    X := V
    Y := V
  }
\end{txnimpcode}
&
\begin{txnimpcode}
  cobegin (i $\leftarrow$ [1$\ldots$3]) {
    txn$\langle$'Inc['+i+']'$\rangle${
      V := V + 1
    }
  }
\end{txnimpcode}
% &
% \begin{txnimpcode}
%   txn$\langle$'Inc2'$\rangle${
%     V := V + 1
%   }
% \end{txnimpcode}
\\
\end{tabular}
\hspace*{-0.3in}$\{\{\texttt{Y}\ge\texttt{X}\}\}$

\caption{View counter (\C{V}) increments (\C{Inc}) and reads (\C{Rd})}
\label{fig:motiv-eg-4}
\end{figure}

Our last example demonstrates that it is not necessary to serialize
transactions, even those that write to the same data objects, to
maintain essential invariants in practical applications. Consider a
counter used to count video views on a YouTube-like application. The
counter need not accurately track the view count, but it must at least
ensure that the count does not decrease (nor does it appear to be
decreasing). Fig.~\ref{fig:motiv-eg-4} shows a simple example
involving a view counter (\C{V}). The three `Inc' transactions
(`Inc[1]' to `Inc[3]') increment the counter, while the transaction
`Rd' reads the counter twice and writes the values read to \C{X} and
\C{Y}, respectively. The count is 0 initially. Since we expect the
view count to monotonically increase, we require $\C{Y}\ge\C{X}$ at
the end. Under the standard setting, where new writes to data objects
always overwrite previous writes, the postcondition may not hold. For
instance, consider an execution where `Inc[1]' reads \C{V} as 0 (but
doesn't commit), then `Inc[2]' and `Inc[3]' commit serially changing
\C{V} to 2, followed by `Rd' reading 2 and writing it to \C{X}. Now
`Inc[1]', which has read \C{V} as 0 writes 1 to \C{V} and commits. The
subsequent read of \C{V} by `Rd' returns 1, which it writes to \C{Y}
and commits. In the final state, \C{X} is 2 and \C{Y} is 1, thus
violating the monotonicity invariant.

However, in an unconventional setting, such as a modern day replicated
data store, writes arriving at a replica need not necessarily
overwrite the existing writes. A \emph{conflict resolution strategy}
may be employed to pick, for example, a write with the largest value
among competing concurrent writes. Under this setting, the write to
\C{V} by `Inc[1]' with a value (1) less than the current value (2) is
discarded. Transaction `Rd' therefore continues to read 2, allowing it
to satisfy the postcondition. Note that we did not have to serialize
`Inc' transactions to uphold the monotonicity invariant, although they
are involved in Write-Write conflicts.  The default \iso{Read
Committed} isolation level suffices\footnote{We have implicitly assumed that
`Rd' transaction witnesses monotonically growing state, i.e., the
transactions that appear committed to the first read of \C{V} in `Rd'
also appear committed to the second read.  Most implementations of RC
enforce this \emph{Monotonic View} invariant. However,
some~\cite{pldi15,bailishat} categorize RC with Monotonic View as a
separate isolation level called \iso{Monotonic Atomic View} (MAV). On
those stores, we need MAV for `Rd'. }, provided that the conflict
resolution strategy consistently picks writes with larger values. A
reasoning framework for weak isolation must be sensitive to the
semantics of conflict resolution in order to formally prove the
monotonicity invariant in this example.

>>>>

In this section, we motivate our proof system for weak isolation
through a series of examples written in \txnimp - a C-like imperative
language equipped with a \C{txn} lexical block for transactions. Each
\C{txn} block is associated with a single-quoted string in angle
braces that uniquely identifies the transaction. We use Hoare triple
notation to annotate programs with pre and post conditions.

@@

% Most weak isolation levels allow executions to interleave operations
% (Reads, Writes and Commits) of one transaction with other's, while
% enforcing some additional constraints. Arbitrary interleavings may
% lead to the violation of invaraints. 
Let us consider \iso{Read Committed} (RC), a weak isolation level
default in Postgres 9.5 and Oracle 11g databases. An RC transaction is
isolated from the writes of an uncommitted transactions, thus
preventing the so called \emph{dirty reads}
phenomenon~\cite{berenson}. In the current example, RC isolation
admits following executions on an SC store, such as a relational
database:

@@

An execution is a series of Read, Write and Commit operations.
Subscript of an operation indicates its transaction. For clarity,
effects of different transactions are shown against differently
colored backgrounds. In the first execution, the first transaction
(`Wd1') reads the current balance (\C{k}) and writes the new balance
(\C{k-a1}), but before it commits transaction `Wd2' executes and
commits, writing the new balance (\C{k-a2}). RC isolation has
prevented the writes of uncommitted transaction `Wd1' from being
\emph{visible} in `Wd2'. Committing `Wd1' now leads to the loss of
updates from `Wd2' (the so called \emph{lost update
anomaly}~\cite{berenson}), resulting in a balance of \C{k-a1}. The
second execution describes a similar scenario with `Wd1' and `Wd2'
exchanging their roles. Clearly, RC is not sufficient because it loses
the updates of one transaction resulting in the violation of post
condition.  We need an isolation level that prevents lost updates.
\iso{Snapshot Isolation} (SI)~\cite{berenson} fits this requirement by
its definition. SI implementations effectively serialize transactions
that update a shared data object mostly by aborting and re-executing a
transaction if write-write conflicts are detected during its commit.
Since SI, unlike SER, need not necessarily rely on expensive
lock-based concurrency control, it is also more efficient, making it
appropriate appropriate for both `Wd' transactions.

Thinking in terms of anomalies, as described above, is how database
programmers are often encouraged to reason about weak isolation.
Unfortunately, such reasoning does not rest on any sound foundation,
thus highly error-prone. Reasoning in terms of the implementations of
weak isolation is not any better since it requires application
programmers to understand and reason about low-level implementation
details of the database (or, databases) that are far removed from the
application semantics. An attractive alternative in this context is a
formal proof system that combines declarative reasoning about
isolation guarantees with operational reasoning about programs. We
demonstrate how our proof system makes this possible in the context of
the current example.

Firstly, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, hence admits \emph{rely-guarantee} style
reasoning~\cite{rgjones}. Rely-guarantee is compositional proof
technique that allows us to reason about one thread at a time by
abstracting away the interference due to remaining threads
(collectively called \emph{the environment}) into a \emph{rely}
relation. In an ordinary concurrent program, every environment step is
a valid interference in the current thread. However, in presence of
transactions and variable isolation, determining what constitutes an
interference and what does not is a non-trivial problem. For example,
inside a serializable transaction no interference is a valid
interference, whereas inside a transaction executing under \iso{Read
Uncommitted}, the weakest isolation level, all interferences are
valid. Between these two extremes are various levels of isolation
that admit some interferences while prohibiting others. 
% For example, \iso{Read Committed} isolation admits interference of a
% committed transaction, but not that of an uncommitted transaction.
% \iso{Snapshot Isolation} admits interference from committed
% non-conflicting transactions, and so on. 
The key aspect of our proof system is that it facilitates
rely-guarantee reasoning while allowing the programmer to ignore any
interference that is invalid as per the chosen isolation level or the
store consistency level. For instance, in the current example, `Wd2'
executes concurrently with `Wd1', hence latter's rely relation is
required to model all possible interferences from `Wd1'. However,
since RC prevents the effects of an uncommitted transaction from being
visible, our proof system allows the programmer to effectively ignore
the interference from `Wd2' unless the interference includes its
commit. RC execution 1 shown above is an example of a case where
interference from `Wd2' includes its commit. Such interference is not
prohibited by RC, and the SC store makes \emph{all} effects of `Wd2'
immediately visible. Our proof system insists that the
programmer consider all effects of `Wd2' and conclude the value of
\C{C} as \C{k-a2}, thus leading (rightfully) to the failure of proof.
On the other hand, \iso{Snapshot Isolation} proscribes any
interference from a concurrent transaction performing conflicting
writes. The RC executions shown above are therefore invalid under SI.
Accordingly, programmer is allowed to ignore any interference from
`Wd2', thus facilitating the proof of postcondition.

As evident from the above example, a proof system for weak isolation
has to be flexible enough to admit different kinds of reasoning for
different isolation and consistency levels. One way this can be
achieved is by having a separate proof rule for each combination of
transaction isolation level and store consistency level, such that the
rule reflects the operational semantics of the isolation level on the
store. Given the large and ever-increasing diversity among isolation
and consistency semantics, this approach is unlikely to scale. Our
proof system avoids this problem by being parametric over the
semantics of isolation and consistency levels. This approach is based
on a couple of observations. First, as noted by some
authors~\cite{pldi15,gotsmanconcur15}, semantics of various isolation
and consistency levels can be captured as well-formedness conditions
over execution traces. If an execution trace satisfies specified
well-formedness conditions, then the program is considered to have
experienced the required levels of consistency and isolation in that
execution. Second, well-formedness conditions are often
\emph{prefix-complete}, meaning that every prefix of a well-formed
trace is also well-formed, or conversely, an ill-formed trace cannot
be extended to a well-formed one. Prefix-completeness allows us to
reason about programs operationally, in terms of an imaginary machine
whose primary artifact is an execution trace. The machine starts from
a well-formed trace, executes a read/write/commit operation, and
extends the trace while preserving well-formedness. If well-formedness
cannot be preserved by taking a step, the machine gets ``stuck''. In
executions that run to completion, the machine models a store whose
consistency and isolation guarantees are captured by the trace
well-formedness conditions. Thus, to show that program's invariants
are preserved under weak isolation, it suffices to show that they are
preserved in all executions of the imaginary machine that do not get
stuck. This is indeed how our proof system facilitates program
verification. The ``invalid interference'' described informally before
corresponds to a case of machine getting stuck due to the violation of
trace well-formedness conditions. The proof system inherits the
parametricity of the machine over trace well-formedness conditions,
leading to a generic set of parameterized proof rules that can be
instantiated for various store consistency and transaction isolation
levels.

We now demonstrate how rely-guarantee reasoning, parametricity ,and
trace well-formedness conditions facilitate the proof of the example
in Fig.~\ref{fig:motiv-eg-1}. The trace-wellformedness condition
corresponding to \iso{Snapshot Isolation} is shown below\footnote{This
is a simplified version of the actual specification of SI
(\S~\ref{sec:opsem}), which is a little more involved.}:
\begin{smathpar}
\begin{array}{lcl}
\underE{\C{SI}(T_j)} & \Rightarrow & \forall T_i. T_i \neq T_j \conj \\
  & & (\exists \C{X}.~\underE{T_i \wrstoar \C{X}} \conj 
                  \underE{T_j \wrstoar \C{X}}) \Rightarrow \\
  &  & \underE{T_i \hboar T_j} \disj \underE{T_j \hboar \C{COMMIT}(T_i)} \\
\end{array}
\end{smathpar}
A transaction $T_j$ that writes to a variable $\C{X}$ has experienced
snapshot isolation in an execution $\E$ only if every other
transaction $T_i$ that also wrote to \C{X} in $\E$ either happened
before (i.e., committed before) $T_j$, or executed concurrently but
commited after $T_j$. Here, $\E$ denotes an execution trace, and
$\underE{\phi}$ asserts that proposition $\phi$ about $\E$ is valid.
Since we require both `Wd1' and `Wd2' to be executed under SI, the
trace invariant ($\I$) for the example is $\C{SI(Wd1)} \wedge
\C{SI(Wd2)}$. That is, if the machine extends an execution $\E$ that
satisfies $\I$ (i.e.,$\underE{\I}$) to $\E'$, then $\E'$ also satisfies
$\I$ (i.e., $\E' \Vdash \I$). Since we reason in terms of traces, pre
and post conditions need to be written with respect to a trace.

\begin{figure}
\centering
\begin{txnimpcode}
 $\begin{decoration}
 P_1:\{ {\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj\\
        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
        \hboar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
 \end{decoration}$
  txn$\langle$'Wd1'$\rangle${
   $\begin{decoration}
    \phi_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj
      {\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \hboar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \}
    \end{decoration}$ 
    v1 = C
   $\begin{decoration}
    \phi_2 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v1 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \hboar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
    \end{decoration}$ 
    if (v1 $\ge$ a1) {
      v2 = C;
     $\begin{decoration}
      \phi_3 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v2 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \hboar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
          \hspace*{1.9in}\C{v2 = k-a2}\}
      \end{decoration}$ 
      v3 = v2 - a1;
     $\begin{decoration}
      \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \hboar \C{Wd1}} 
         \Rightarrow \C{C = k-a2} \wedge \\
         \hspace*{1.9in}\C{v3 = k-a2-a1}\}
      \end{decoration}$ 
      C := v3
     $\begin{decoration}
      \phi_5 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
                \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} 
                \Rightarrow \C{C = k-a1-a2}\}
      \end{decoration}$ 
    }
  }
 $\begin{decoration}
  Q_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}
            \conj \committed(\C{Wd1}) \conj\\
      \hspace*{0.12in}{\committed(\C{Wd2})} 
          \Rightarrow \C{C = k-a1-a2} \}
  \end{decoration}$ 
\end{txnimpcode}

\caption{`Wd1' transaction decorated with assertions}
\label{fig:wd1-decorated}
\end{figure}

The fully decorated implementation of transaction `Wd1' is shown in
Fig.~\ref{fig:wd1-decorated}. Decorated `Wd2' is similar. Assignment
statements are broken down and temporary local variables (\C{v1},
\C{v2} and \C{v3}) are introduced so as to separate shared variable
reads and writes.  Since we reason in terms of executions, all
assertions implicitly refer to the current execution (just as hoare
triples implicitly refer to the current state). Also implicit is the
invariant \C{k $\ge$ a1+a2}. Proposition $\committed(T)$ indicates
that the transaction $T$ is already committed. Precondition ($P_1$) of
`Wd1' accounts for the possibility of `Wd2' committing before `Wd1',
writing \C{k-a2} to \C{C}.  Precondition of `Wd2' is similar.  Since
neither `Wd1' nor `Wd2' are committed at the beginning, the
precondition in Fig.~\ref{fig:motiv-eg-1} is extended to include
$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
$P_1$ follows. Once the execution is inside `Wd1', commit of `Wd2'
($\committed(\C{Wd2})$) may mean that either `Wd2' happened before
`Wd1', or that it committed concurrently with `Wd1'. Since SI
proscribes the latter possibility, we only consider the case when
$\C{Wd2} \hboar \C{Wd1}$. A proof for $\C{Wd2} \hboar \C{Wd1}$ is
obtained subsequently, allowing us to get rid of the special case.

The first obligation is to show that the assertions that decorate
`Wd1' are indeed valid. This entails proving that, for each statement
of `Wd1', if evaluating the statement and extending an execution that
satisfies its precondition results in a well-formed execution, then
the extended execution must satisfy the postcondition. The
well-formedness assumption lets us ignore the executions that violate
the trace invariant lead the machine to a stuck state. This is useful
in case of the assignment statement that writes \C{v3} to \C{C}. Since
`Wd2' also writes to \C{C}, the trace invariant ($\I$) requires
$\C{Wd2} \hboar \C{Wd1}$ if $\committed(\C{Wd2})$ (the alternative
$\C{Wd1} \hboar \C{COMMIT}(\C{Wd2})$ allowed by \C{SI(Wd1)} is
impossible in this case because \C{Wd2} has already committed, while
\C{Wd1} is still in progress). This lets us treat the special case
$\committed(\C{Wd2})\conj \C{Wd2} \hboar \C{Wd1}$ as being equivalent
to the general case $\committed(\C{Wd2})$, which is required to prove
the postcondition of `Wd1'. Observe that \C{RC(Wd1)} does not justify
this generalization, hence replacing \C{SI(Wd1)} by \C{RC(Wd1)} is not
sufficient to prove the postcondition of `Wd1'.

The second obligation is to prove that assertions are valid despite
the interference from the the concurrent thread executing `Wd2'. The
interference is given by the following rely relation ($R_1$):
\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
        \underE{\I} \conj \E'\Vdash \I \conj\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
\end{array}
\end{smathpar}
The rely relation says the following about the concurrent thread: (1).
It may interfere only if `Wd2' is not already committed, (2). Any
interference takes well-formed executions to well-formed executions,
(3). If the interference commits `Wd2', then `Wd2' should have already
written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
or $\C{k-a1}$ depending on whether or not `Wd1' has already committed.
Interference from `Wd2' is allowed between any two operations of
`Wd1', and assertions are required to be valid despite the
interference. For instance, if the execution ($\E$) immediately after
evaluating $\C{v1 = C}$ satisfies $\phi_2$ (i.e., $\underE{\phi_2}$),
and an interference from `Wd2' extends $\E$ to $\E'$ (i.e.,
$R_1(\E,\E')$), then the execution $\E'$ just before evaluating the
condition must satisfy $\phi_2$ (i.e., $\E' \Vdash \phi_2$). Indeed,
all $\phi_i$'s do satisfy this \emph{stability} condition, thus
lending the validity to the proof.
